Treebeard : An Optimizing Compiler for Decision Tree Based ML Inference

### 2022-04-25 docs/MICRO 2022/related.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\section{Related Work}
While decision trees are heavily used there is little prior work on building optimizing compilers for tree 
based inference. Below we discuss some related work and contrast it to \Treebeard{}.

\emph{Decision Tree Ensemble Compilers:}
Treelite\cite{Treelite} is a model compiler for decision tree ensembles and is the system
most closely related to our work. Even though Treelite compiles ensembles, it only supports 
generation of simple \op{if-else} style code. It does not perform the rich 
optimizations \Treebeard{} performs and is not easily extensible.
% It cannot perform any of the optimizations \Treebeard{} is designed to perform. 
Hummingbird\cite{Hummingbird} compiles traditional ML models to make use 
of tensor primitives so 
they can be integrated into tensor based frameworks like TensorFlow \cite{TensorFlow}.
Hummingbird does not perform any optimizations tailored to decision trees. It 
also does not perform model specific optimizations.

\emph{Libraries:} Currently, the most popular systems for decision tree based models 
are libraries. XGBoost\cite{XGBoost} and LightGBM\cite{LightGBM} are the most  
popular gradient boosting libraries while scikit-learn\cite{Sklearn} is 
extremely popular for random forest models. These libraries implement both 
training and inference. However, as mentioned in section \ref{sec:Intro}, porting 
optimizations in these libraries to newer architectures and maintaining them is 
extremely difficult. Additionally, the inference routines in these libraries have 
to be general and cannot be specialized to a specific model.

@@

\emph{Other Systems and Techniques}: 
Ren et al \cite{PortableVM} build an intermediate language and VM to 
enable SIMD execution of decision tree inference. The SIMD execution itself is implemented 
by hand in the VM and the VM needs to be reimplemented for every supported target architecture.
Additionally, even though they perform layout optimizations, their system does not perform 
any model specific optimizations. Jo et. al.\cite{MilindTreeVectorization} describe techniques 
to vectorize tree based applications. They do not study optimizations specific to decision trees.
Both these works vectorize tree walks by performing different tree walks on each vector lane.
The main issue with this approach is the divergence of tree walks. Another issue is that 
memory accesses are gathers rather than vector loads. \Treebeard{}'s approach to 
vectorization solves both these issues. Also, these approaches are not precluded 
by \Treebeard{}'s tiling based vectorization. Multiple tiled tree walks can be combined
into a single vectorized walk. We leave an exploration of this to future work. 
FAST\cite{FAST} is a system that accelerates tree structured index search 
on CPUs and GPUs. FAST defines a layout for the index tree that enables vectorization 
of the tree walk. FAST uses a tree tiling approach to vectorize tree walks. However, 
FAST only uses a single triangular tile shape. \Treebeard{}'s basic tiling algorithm 
is a generalization of the tiling used in FAST. If given a perfectly balanced tree,
the basic tiling algorithm would return exactly the tiling used by FAST. 
Also, the tree walks in FAST are hand coded using intrinsics on the CPU and CUDA 
on the GPU and therefore need repeated effort to implement on each target.

\emph{Code Generation Systems from Other Domains:}
Several compilers and code generation techniques exist for other domains.
TVM\cite{TVM}, Tiramisu\cite{Tiramisu} and Tensor 
Comprehensions\cite{TensorComprehensions} are domain specific compilers 
for deep learning models. Halide\cite{Halide} is a DSL and compiler 
primarily designed for image processing applications. Several systems
that generate optimized processing routines have also been designed 
for BLAS and signal processing. BLIS\cite{BLIS} and ATLAS\cite{atlas_sc98}
are systems to instantiate high performance BLAS routines on multiple 
target architectures. CUTLASS\cite{CUTLASS} provides building blocks 
in the form of C++ templates to quickly instantiate high performance 
BLAS functions on different GPU architectures. SPIRAL\cite{SPIRAL}
is a domain specific compilers for signal processing applications
that instantiates high performance routines for linear transformations 
like FFTs and FIR filters. FFTW\cite{FFTW} is a fast fourier transform 
compiler that generates high performance FFT routines by customizing 
the routine based on FFT size and the target machine.

>>>>

\section{Related Work}
While decision trees are heavily used there is little prior work on building 
optimizing compilers for tree-based inference. We discuss below related work and 
contrast it to \Treebeard{}.

\emph{Decision Tree Ensemble Compilers:}
Treelite\cite{Treelite} is a model compiler for decision tree ensembles and is the system
most closely related to our work. Even though Treelite compiles ensembles, it only supports 
generation of simple \op{if-else} style code. It does not perform the rich 
optimizations \Treebeard{} performs and is not easily extensible.
% It cannot perform any of the optimizations \Treebeard{} is designed to perform. 
Hummingbird\cite{Hummingbird} compiles traditional ML models to make use 
of tensor primitives so 
they can be integrated into tensor-based frameworks like TensorFlow \cite{TensorFlow}.
Hummingbird does not perform any optimizations tailored to decision trees. It 
also does not perform model specific optimizations.

\emph{Libraries:} Currently, the most popular systems for decision tree-based models 
are libraries. XGBoost\cite{XGBoost} and LightGBM\cite{LightGBM} are the most  
popular gradient boosting libraries while scikit-learn\cite{Sklearn} is 
extremely popular for random forest models. These libraries implement both 
training and inference. However, as mentioned in Section \ref{sec:intro}, porting 
optimizations in these libraries to newer architectures and maintaining them is 
extremely difficult. Additionally, the inference routines in these libraries have 
to be general and cannot be specialized to a specific model.

@@

\emph{Other Systems and Techniques}: 
Ren et al \cite{PortableVM} build an intermediate language and VM to 
enable SIMD execution of decision tree inference. The SIMD execution itself is implemented 
by hand in the VM and the VM needs to be reimplemented for every supported target architecture.
Additionally, even though they perform layout optimizations, their system does not perform 
any model specific optimizations. Jo et. al.\cite{MilindTreeVectorization} describe techniques 
to vectorize tree-based applications. They do not study optimizations specific to decision trees.
Both these works vectorize tree walks by performing different tree walks on each vector lane.
The main issue with this approach is the divergence of tree walks. Another issue is that 
memory accesses are \op{gather}'s rather than vector loads. \Treebeard{}'s approach to 
vectorization solves both these issues. Also, these approaches are not precluded 
by \Treebeard{}'s tiling based vectorization. Multiple tiled tree walks can be combined
into a single vectorized walk. We leave an exploration of this to future work. 
FAST\cite{FAST} is a system that accelerates tree structured index search 
on CPUs and GPUs. FAST defines a layout for the index tree that enables vectorization 
of the tree walk. FAST uses a tree tiling approach to vectorize tree walks. However, 
FAST only uses a single triangular tile shape. \Treebeard{}'s basic tiling algorithm 
is a generalization of the tiling used in FAST. If given a perfectly balanced tree,
the basic tiling algorithm would return exactly the tiling used by FAST. 
Also, the tree walks in FAST are hand coded using intrinsics on the CPU and CUDA 
on the GPU and therefore need repeated effort to implement on each target.

\emph{Code Generation Systems from Other Domains:}
Several compilers and code generation techniques exist for other domains.
TVM\cite{TVM}, Tiramisu\cite{Tiramisu} and Tensor 
Comprehensions\cite{TensorComprehensions} are domain specific compilers 
for deep learning models. Halide\cite{Halide} is a DSL and compiler 
primarily designed for image processing applications. Several systems
that generate optimized processing routines have also been designed 
for BLAS and signal processing. BLIS\cite{BLIS} and ATLAS\cite{atlas_sc98}
are systems to instantiate high performance BLAS routines on multiple 
target architectures. CUTLASS\cite{CUTLASS} provides building blocks 
in the form of C++ templates to quickly instantiate high performance 
BLAS functions on different GPU architectures. SPIRAL\cite{SPIRAL}
is a domain specific compilers for signal processing applications
that instantiates high performance routines for linear transformations 
like FFTs and FIR filters. FFTW\cite{FFTW} is a fast fourier transform 
compiler that generates high performance FFT routines by customizing 
the routine based on FFT size and the target machine. 
However, no such frameworks for decision tree ensembles exist currently.
\Treebeard{} is a first step in this direction and unlocks several future 
optimization opportunities.

### 2022-04-25 docs/MICRO 2022/tiling.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\subsection{Tiling}
\label{sec:Tiling}
% Treebeard vectorizes tree walks by grouping nodes of a decision tree into \textbf{\emph{tiles}}. The nodes in a tile are evaluated concurrently using vector instructions. Once the nodes of the current tile are evaluated, a look up table is used to compute which child of the current tile to move to next.

While a decision tree is naturally represented by a binary tree, 
it is not the best representation for tree traversal as it (i) requires many memory accesses, 
(ii) has poor branching structure and (iii) cannot make use of vector instructions. 
This section proposes a tiling optimization where we group multiple nodes in a decision tree into a single tile, effectively transforming a binary tree into an $n$-ary tiled tree. This not only allows the compiler to generate vectorized code (see Section~\ref{sec:Vectorization} ) to traverse trees but also enables spatial locality improvements by grouping nodes that are likely to be accessed together. We demonstrate this with two tiling heuristics later in the section.

@@

Once trees are tiled Treebeard generates tree walks with the code structure shown below.
\begin{lstlisting}[style=c++]
  ResultType Prediction_Function(...) {
    // ...
    Tile t = getRootTile(tree)
    while (!isLeaf(tree, t)) do {
      // Evaluate predicates of all nodes in the tile
      predicates = evaluateTilePredicates(t, rows[i])

      // Move to the correct child of the current tile
      t = getChildTile(tree, t, predicates) 
    }
    treePrediction = getLeafValue(t)
    // ...
  }  
\end{lstlisting}
The code is just an abstract representation of a tiled tree walk that enables efficient lowering of specific steps in subsequent stages. 
\op{evaluateTilePredicates} (speculatively) computes the predicates on all nodes in a tile (line 6). Then \op{getChildTile} (line 9), uses the computed predicate values to determine which child of the current tile to move to. We defer a description of how these operators are lowered to a later section but focus on tiling algorithms in this section.

@@

\subsection{Tiles and Tree Tiling}
\label{sec:ValidTiling}
}
\subsubsection*{Conditions for Efficient Tiling}
\label{sec:ValidTiling}
While any arbitrary partitioning of the nodes of a tree could be considered for tiling we impose a few intuitive constraints.
% to only allow nodes that are likely to be accessed together to be grouped into a tile. 
% \TODO{kr : replace $n_t$ with sz?}
Given a tree $\Tree = (V, E, r)$ and a tile size $n_t$ we impose the following constraints on the generated tiles $\{ T_1, T_2, ... ,T_m \}$ .
\begin{description}
    \item[Partitioning] $T_1 \cup T_2 ... \cup T_m = V$ and $T_i \cap T_j = \emptyset$ for all $i\neq j$
    \item[Connectedness] If $u, v \in T_i$, there is a (undirected) path connecting $u$ and $v$ fully contained in $T_i$.
   \item [Leaf seperation] $\forall l \in L_{\Tree}$ : $l \in T_i \rightarrow v \notin T_i \;\; \forall v \in V \backslash \{l\}$
  \item [Maximal tiling] if there are tiles such that. $|T_i| < n_t$ then there is no $v \in V\backslash \{ T_i \cup L_{\Tree} \}$ such that $(u, v) \in E$ for some $u \in T_i$. 
\end{description}
The \textbf{partitioning} and \textbf{maximal tiling} constraints together ensure that we group nodes into as few tiles as possible. {\textbf{Connectedness}} ensures that each tile is a sub-tree, a natural grouping of nodes that are likely to be accessed together. The {\textbf{Leaf separation}} constraint ensures that leafs are not tiled along with internal nodes. Leafs in a decision tree need special handling, they are used to check for walk termination and to determine the output (prediction). This constraint ensures that tiles are homogenous, this in-turn allows us to specialize the in-memory layout of trees and also simplifies code generation. We discuss leaf handling and tree layout in section ~\ref{Sec:MemoryRep}. We refer to any tiling that satisfies the above constraints as a \emph{valid} tilling.

@@

\input{uniftiling}
\input{probtiling}

\subsection{A Note on Implementation}
\TODO{Kr : Not sure if this is needed}
The tiling algorithms generate a \op{TileId} attribute per tree. The \op{TileId} attribute contains a mapping from a Node to the TileId asigned to it.
This information is used when lowering to the mid level abstraction in the form of loops. A sample of the lowered MIR code is shown in Figure~\ref{Fig:Overview}.

>>>>

\subsection{Tree Tiling}
\label{sec:Tiling}
% Treebeard vectorizes tree walks by grouping nodes of a decision tree into \textbf{\emph{tiles}}. The nodes in a tile are evaluated concurrently using vector instructions. Once the nodes of the current tile are evaluated, a look up table is used to compute which child of the current tile to move to next.

While a decision tree is naturally represented as a binary tree, 
this is not an efficient representation for tree traversal as it (i) requires many memory accesses, 
(ii) has poor branching structure which may cause high branch misprediction rates and (iii) cannot use vector instructions. 
This section proposes a tiling optimization where multiple nodes of a decision tree are grouped into a single tile,
effectively transforming a binary tree into an $n$-ary tiled tree. 
Note that an ideal decision tree traversal visits only a subset of nodes in a tile, with strict dependency between nodes at different levels.
Unfortunately, this dependency prevents parallelization/vectorization of the tree walk. However, by speculatively evaluating conditions 
of all nodes in a tile, we can eliminate this dependency and enable
the compiler to generate vectorized code (see Section~\ref{sec:Vectorization}) to traverse trees.
Secondly, this enables spatial locality improvements by grouping together nodes that are likely to be accessed together (Section~\ref{Sec:MemoryRep}).   
We present two different tiling heuristics later in this section.

@@

Once trees are tiled \Treebeard{} generates tree walks with the code structure shown below.
\begin{lstlisting}[style=c++]
  WalkDecisionTree(...) {
    tile = getRootTile(tree)
    while (!isLeaf(tree, tile)) do {
      // Evaluate predicates of all nodes in the tile
      predicates = evaluateTilePredicates(tile, rows[i])

      // Move to the correct child of the current tile
      tile = moveToChildTile(tree, tile, predicates) 
    }
    treePrediction = getLeafValue(tile)
  }  
\end{lstlisting}
The code is an abstract representation of a tiled tree walk that enables efficient lowering of specific steps in subsequent stages. 
\op{evaluateTilePredicates} (speculatively) computes the predicates of all nodes in a tile (line 6)\footnote{ 
\op{evaluateTilePredicates} expands to the operations from line 10 to 15 in the listing in Section~\ref{sec:Vectorization}.}.
Then \op{moveToChildTile} (line 9), 
uses the computed predicate values to determine which child of the current tile to move to\footnote{\op{moveToChildTile} expands 
to the operations from line 18 to 25 in the listing in Section~\ref{sec:Vectorization}.}. We defer a description of how these 
operators are lowered to Section~\ref{sec:Vectorization} and instead focus on tiling algorithms in this section.

@@

\subsubsection{Tiles and Tree Tiling}
\label{sec:ValidTiling}
}
\subsubsection{Conditions for Valid Tiling}
\label{sec:ValidTiling}
While any arbitrary partitioning of the nodes of a tree could be considered for tiling, we impose a 
few simple constraints to simplify the design of the compiler.
% to only allow nodes that are likely to be accessed together to be grouped into a tile. 
% \TODO{kr : replace $n_t$ with sz?}
Given a tree $\Tree = (V, E, r)$ and a tile size $n_t$ we impose the following constraints on the generated tiles $\{ T_1, T_2, ... ,T_m \}$.
\begin{description}
    \item[Partitioning:] $T_1 \cup T_2 ... \cup T_m = V$ and $T_i \cap T_j = 
    \emptyset$ for all $i\neq j$.
    \item[Connectedness:] If $u, v \in T_i$, there is a (undirected) path connecting $u$ and $v$ fully contained in $T_i$.
   \item [Leaf separation:] $\forall l \in L_{\Tree}$ : $l \in T_i \rightarrow v \notin T_i \;\; \forall v \in V \backslash \{l\}$.
  \item [Maximal tiling:] if there are tiles such that $|T_i| < n_t$, then there 
is no $v \in V\backslash \{ T_i \cup L_{\Tree} \}$ such that $(u, v) \in E$ for 
some $u \in T_i$.  \end{description}
The \textbf{partitioning} and \textbf{maximal tiling} constraints together ensure that we group nodes into as few tiles as possible. {\textbf{Connectedness}} ensures that each tile is a sub-tree, a natural grouping of nodes that are likely to be accessed together. The {\textbf{leaf separation}} constraint ensures that leaves are not grouped with internal nodes, as leaves in a decision tree need special handling and are used to check for walk termination and to determine the output (prediction). Thus, the {\textbf{leaf separation}} constraint ensures that tiles are homogeneous. This in-turn allows \Treebeard{}  to specialize the in-memory layout of trees and also simplifies code generation. We discuss leaf handling and tree layout in Section ~\ref{Sec:MemoryRep}. We refer to any tiling that satisfies the above constraints as a \emph{valid} tiling.

@@

\subsubsection{Rationale for Different Tiling Methods}
\label{sec:rationale}

Before we present different tiling methods, we first discuss an interesting property of decision trees and 
explore whether it can be exploited to optimize tree traversal. We find that the probabilities of reaching different leaves
in a tree can vary significantly, making some root to leaf paths more frequently traversed than others. As the set of 
decision trees that are walked for a given ML model is fixed, we ask whether this statistical property can be 
exploited to generate efficient traversal code.% We address this question by developing different decision tree tiling methods.

To understand how often different leaves are reached, let us look at the percentage of leaves required to 
cover a significant part (say 80\% or 90\%) of the training data\footnote{We use the training data to compute tree statistics 
since the test data must have a distribution similar to the training data.}.
% in the decision tree ensemble. 
This is captured in Figures~\ref{Fig:AirlineOHEStats} and~\ref{Fig:EpsilonStats}.
%% that are generated from training data. 
Each line in these graphs corresponds to a fixed fraction  (say $f$) of the training data. 
A point on this line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a
fraction $f$ of all training inputs with a fraction $x$ of leaves. For example, the first point on the
$f=0.9$ line in Figure \ref{Fig:AirlineOHEStats}(at the bottom left of the plot) indicates that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure~\ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction 
of inputs for the benchmark \op{airline-ohe}. This means that a small fraction of leaves are very likely. 
We call trees with a small number of extremely likely leaves \textbf{\emph{leaf-biased}}.
On the other hand, for the benchmark \op{epsilon},
Figure \ref{Fig:EpsilonStats} shows that trees need a much larger fraction of their leaves to cover a significant fraction of the training input.
This means that most trees in \op{epsilon} are not leaf-biased. For leaf-biased trees, it would be beneficial to tile nodes such that the
depth of the most probable leaves are minimized even at the expense of 
increasing the depth of the less probable leaves. We call the tiling algorithm 
designed to do this \emph{probability-based tiling} (Section \ref{sec:ProbTiling}).

% \TODO{AP One hole here is why doesn't this method of tiling work for unbiased trees.}

\begin{figure*}[tbp]
  \centering
  \subfloat[\texttt{airline-ohe}]{
    \centering
    \includegraphics[width=0.35\linewidth]{figures/airline-ohe.png}
    \label{Fig:AirlineOHEStats}
  }
  %\newline
  \hspace{2cm}
  \subfloat[\texttt{epsilon}]{
    \centering
    \includegraphics[width=0.35\linewidth]{figures/epsilon.png}
    \label{Fig:EpsilonStats}
  }
  \caption{Statistical profiles for \texttt{airline-ohe} and \texttt{epsilon}.}
\end{figure*}

% From the above discussion, we can infer that when the decision tree ensemble is not leaf-biased, it would be beneficial to tile the trees in 
% such a way that the depth of the tree (in terms of the number of tiles) is minimized across all the leaves (assuming there is no major skew 
% in the probabilities of different leaves).

However, when there is no clear leaf-bias, a reasonable objective is to minimize the number of nodes executed 
speculatively. An intuitive heuristic to achieve this objective is to minimize the depth of each tile 
that is constructed. \emph{Basic tiling} (Section \ref{sec:UnifTiling}) is a heuristic designed to do this. 
It is possible to define other variants of tiling(e.g., minimize the maximum leaf depth), but we leave this for future work.
We believe that \Treebeard{}'s tiling infrastructure can support and optimize all such variants that produce valid tilings.

\CommentOut{
\subsubsection{The Optimization Problem}

%We assume that we are given the probabilities of each leaf node of the decision tree (these can easily be computed using the training data). For every leaf $l \in L$, we %are given the probability $p_l$ that the leaf $l$ is reached.

Observe that the latency of one tree walk is proportional to the number of tiles that need to be evaluated to reach the leaf. It is easy to see that for a leaf biased tree, basic tilling does not optimize for this objective, it considers all leaves to be equally likely.

The goal of probablistic tiling is to minimize the average inference latency, or equivalently the minimize the expected number of tiles that are evaluated to compute one tree prediction. More formally, the problem is to find a \emph{valid} (as defined in Section~\ref{sec:ValidTiling}) tiling $\mathcal{T}$ such that the following objective is minimized.
\[
    \min_{\mathcal{T} \in \mathcal{C}(T)}{\sum_{l \in L_{\Tree}} p_l.depth_{\mathcal{T}}(l)}
\]
where the minimization is over all valid tilings $\mathcal{T}$ of the tree $\Tree$, $depth_{\mathcal{T}}(l)$ is the depth of the leaf $l$ given tiling ${\mathcal{T}}$. $p_l$ is the probability of of reaching leaf $l$ as observed during training.

The above optimization problem can be solved optimally using dynamic programming. 
We leave this out in the interest of space. 
Instead, we use the simple greedy algorithm listed in algorithm \ref{Alg:GreedyTilingAlgo} to construct a valid tiling given the node probabilities\footnote{Probabilites for internal nodes can be computed from probablities for leaves by summing up the probabilities of all leaves that belong to the sub-tree rooted at the internal node. Leaf probabilities are collected during training.}.
The algorithm starts at the root and greedily keeps adding the most probable legal node to the current tile until the maximum tile size is reached.
Subsequently, the tiling procedure is recursively performed on all nodes that are destinations for edges going out of the constructed tile.
}

\input{probtiling}
\input{uniftiling}

\subsection{A Note on Implementation}
%\TODO{Kr : Not sure if this is needed}
The tiling algorithms generate a \op{TileId} attribute per tree. The \op{TileId} attribute 
contains a mapping from a Node to the TileId assigned to it.
This information is used when lowering to the mid level abstraction in the form of loops. 
%A sample of the lowered MIR code is shown in Figure~\ref{Fig:Overview}.

### 2022-04-25 docs/MICRO 2022/treeorder.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\subsection{Tree ordering}
\label{sec:treeorder}	
%Another objective at the HIR is to reduce the number of trees that need to be specialized by the lower levels.   
% \TODO{kr : Not sure if i got this para right}
Specializing the code for each tree in a model comes at a cost. First, the size of the generated code increases if the
code generator needs to generate different code for different trees. This causes several front end stalls like instruction 
cache misses and delays in instruction decoding. Second, some cross tree optimizations 
like tree walk interleaving (applied at the lower levels of abstraction) are more effective when multiple trees share identical code.

In order to handle this, \Treebeard{} groups trees by their structure so that trees with the same structure can 
share code. For example, the compiler pads trees with dummy tiles to make them balanced and then sorts the trees by 
their depth, so that trees of same depth can share the same unrolled tree walk. Padding is only done for almost balanced trees (as generated 
by basic tiling). This is shown in the example in figure \ref{Fig:Overview}. Other metrics, like feature set commonality, 
could also be used to reorder trees\cite{Tahoe}. However, we leave the exploration of such optimizations to future work.

>>>>

\subsection{Tree Reordering}
\label{sec:treeorder}	
%Another objective at the HIR is to reduce the number of trees that need to be specialized by the lower levels.   
% \TODO{kr : Not sure if i got this para right}
Specializing the code, as we do later in MIR for example by unrolling tree walks, for each tree in a model comes at a cost.
First, the size of the generated code increases if the
code generator needs to generate different code for different trees. This causes several front end stalls like instruction 
cache misses and delays in instruction decoding. Second, some cross tree optimizations 
like tree walk interleaving (applied at the lower levels of abstraction) are more effective when multiple trees share identical code.

In order to handle this, \Treebeard{} groups trees by tree structure so that they can 
share traversal code. For example, the compiler pads trees with dummy tiles to make them balanced and then sorts the trees by 
their depth, so that isomorphic trees can share the same unrolled tree walk. Padding is only done for almost balanced trees (as generated 
by basic tiling). Additionally, the loop rewriting infrastructure described in Section~\ref{Sec:LoopReordering} is used 
to modify the loop nest so that the same code walks all trees in a group. 
This is shown in the example in the third row of Figure \ref{Fig:Overview}. Other metrics, like feature set commonality, 
could also be used to reorder trees\cite{Tahoe}. However, we leave the exploration of such optimizations to future work.

### 2022-04-25 docs/MICRO 2022/uniftiling.tex
Micro paper (#37)
* Add results for multiple batch sizes run

* Changes to benchmark code

* Add xgboost and treelite compare results

* Add AMD results

* Add parallel performance results

* code to generate graphs + modified result numbers

* Fix bug in plotting code

* Add grid to line graphs

* Add grid lines

* Add results for varying number of cores

* Run parallel tests

* Add AMD parallel results

* Update graph generator

* Add scaling results for AMD

* Fix images, alter font sizes and axes

* Just use Probability based tiling

* Update images - Change axis font size, fix wrong graph

\subsection{Basic Tiling Algorithm}
\label{sec:UnifTiling}
% The justification for basic tiling -- if we assume all paths are equally likely,
% we want to minimize the depth of speculation
Algorithm~\ref{Alg:UnifTilingAlgo} shows a simple tiling algorithm that satisfies the above constraints and produces a valid tiling. 
Tiling starts at the root and constructs a tile $Tile$ by performing
a level order traversal. The call \op{LevelOrderTraversal($\Tree$, $n_t$)} picks the next $n_t$ non-leaf nodes according to the standard 
level order tree traversal algorithm. Once the current tile is constructed, 
the tiling procedure is recursively performed on subtrees rooted at each node that is a destination of an edge going out of the constructed 
tile. These edges are denoted by $Out(Tile)$ and the subtree rooted at a node $v$ is denoted by $S_v$.
It is easy to see that the tiling constructed by algorithm \ref{Alg:UnifTilingAlgo} is valid.
% \TODO{Pass a modified tree without leafs to the level order traversal call. Explicitly add leaves as seperate tiles}
% Algorithm for basic tiling
\begin{algorithm}
  \caption{Basic tree tiling}
  \label{Alg:UnifTilingAlgo}
  \begin{algorithmic}
      \Procedure{LevelOrderTraversal}{$\Tree = (V, E, r)$, $n_t$}
        \State $queue = \{ r \}$
        \State $Tile = \emptyset$
        \While{$\neg queue.empty() \wedge |Tile|<n_t$}
          \State $n = queue.pop()$
          \If {$n \in L_{\Tree}$}
              \State \textbf{continue}
          \EndIf
          \State $Tile = Tile \cup \{ n \}$
          \State $queue.push(left(n), right(n))$
        \EndWhile
      \EndProcedure
      \State {}
      \Procedure{TileTree}{$\Tree = (V, E, r)$, $n_t$} 
          \If {$r \in L_{\Tree}$}
              \State \textbf{return} $\{ r \}$
          \EndIf
          % \State \textcolor{codegreen}{\textit{//Level order traversal to collect $n_t$ or fewer nodes. }}
          % \State \textcolor{codegreen}{\textit{//Leaves are not included in the constructed tile. }}
          \State $Tile \leftarrow LevelOrderTraveral(\Tree, n_t)$
          \State $Tiles =  \{ Tile \}$
          \For{$(u,v) \in Out(Tile)$}
              \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
          \EndFor
          \State \textbf{return} $Tiles$
      \EndProcedure
  \end{algorithmic}
\end{algorithm}

One interesting property of this tilling algorithm is that it naturally reduces the imbalance in trees, especially at large tile sizes. As the algorithm traverses down to sparser levels of the tree, it naturally groups sub-trees containing chains of nodes, thus balancing the trees. While its possible to further enhance the algorithm to explicitly balance tiled trees, we find that basic tiling suffices in practice.

>>>>

\subsection{Basic Tiling}
\label{sec:UnifTiling}
% The justification for basic tiling -- if we assume all paths are equally likely,
% we want to minimize the depth of speculation
Algorithm~\ref{Alg:UnifTilingAlgo} shows the basic tiling algorithm that produces a valid tiling. It 
attempts to minimize the depths of all constructed tiles. 
Tiling starts at the root and constructs a tile $Tile$ by performing
a level order traversal. The call \op{LevelOrderTraversal($\Tree$, $n_t$)} picks the next $n_t$ non-leaf nodes according to the standard 
level order tree traversal algorithm. Once the current tile is constructed, 
the tiling procedure is recursively performed on subtrees rooted at each node that is a destination of an edge going out of the constructed 
tile. It is easy to see that the tiling constructed by Algorithm \ref{Alg:UnifTilingAlgo} is valid.
% \TODO{Pass a modified tree without leaves to the level order traversal call. Explicitly add leaves as seperate tiles}
% Algorithm for basic tiling
\begin{algorithm}
  \caption{Basic tree tiling}
  \label{Alg:UnifTilingAlgo}
  \small{
  \begin{algorithmic}[1]
      \Procedure{LevelOrderTraversal}{$\Tree = (V, E, r)$, $n_t$}
        \State $queue = \{ r \}$
        \State $Tile = \emptyset$
        \While{$\neg queue.empty() \wedge |Tile|<n_t$}
          \State $n = queue.dequeue()$
          \If {$n \in L_{\Tree}$}
              \State \textbf{continue}
          \EndIf
          \State $Tile = Tile \cup \{ n \}$
          \State $queue.enqueue([left(n), right(n)])$
        \EndWhile
      \EndProcedure
      \State {}
      \Procedure{TileTree}{$\Tree = (V, E, r)$, $n_t$} 
          \If {$r \in L_{\Tree}$}
              \State \textbf{return} $\{ r \}$
          \EndIf
          % \State \textcolor{codegreen}{\textit{//Level order traversal to collect $n_t$ or fewer nodes. }}
          % \State \textcolor{codegreen}{\textit{//Leaves are not included in the constructed tile. }}
          \State $Tile \leftarrow LevelOrderTraveral(\Tree, n_t)$
          \State $Tiles =  \{ Tile \}$
          \For{$(u,v) \in Out(Tile)$}
              \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
          \EndFor
          \State \textbf{return} $Tiles$
      \EndProcedure
  \end{algorithmic}
  }
\end{algorithm}

One interesting property of this tiling algorithm is that it naturally reduces 
the imbalance in trees, especially at large tile sizes. As the algorithm 
traverses down to sparser levels of the tree, it naturally groups sub-trees 
containing chains of nodes, thus balancing the trees. While it is possible to 
further enhance the algorithm to explicitly balance tiled trees, we find that 
basic tiling suffices in practice.
